Proportional multi-state multiple-cohort life table model

### 2018-04-05 MSLT/MSLTdoc.Rmd
Inputs summary table

---
title: "Proportional multi-state multiple-cohort life table model"
author: "Belen Zapata-Diomedi and Ali Abbas"
date: "26 March 2018"
header-includes:
    - \usepackage{setspace}\onehalfspacing
    - \usepackage{float}
output:
   pdf_document: 
     toc: true
     number_sections: true
     fig_caption: yes

   fontsize: 12 pt
   geometry: margin=1in
bibliography: Bibliography.bib
---

@@

![Figure 1. Basic model infrastructure.](structure\Figure1.png)

@@

![Figure 2. Proportional multi-state life-table simplified framework. *The simplied PMST shows the interaction between the life table, disease life table and potential impact fraction (PIF). The PIF calculations by age and sex group are the same as those generated for ITHIM. The PIF (or 1-PIF) modifies incidence of disease, which changes prevalence and mortality (disease specific life table). Changes in prevalence and mortality rates from the disease specific life tables feed into the life table by changing all-cause mortality, which in turn changes life years. Change in prevalence of diseases changes total years lived with disability, which in turn modifies health-adjusted life years*](structure\Figure2.png)

@@

### **Life table**:

Inputs of the life table are: population numbers by sex (per 1-year or age grouping of interest), mortality rates or probability of all cause mortality by single age group and sex and total prevalent years lived with disability rate per one year by sex.

@@

Data source: (1) National census; (2) Worldwide population and mortality data:  http://www.mortality.org/ (mostly high income countries; and (3) Calculate from GBD data (rates and numbers).

@@

These data will be provided by the synthetic population. Other sources are the same sources as above for population numbers. Sometimes mortality rates are in age groups (1-4, 5-9, etc). Interpolation can be used to derive in between ages rates (cubic spline). Here, mortality rates are needed per single year and sex.

Note that we need data for population numbers and all cause mortality rates for: (1) PMSLT and (2) Dismod II collection (more in Dismod II section).

@@

### **Disease life tables**

@@

For each of the modeled diseases the PMSLT needs incidence and case fatality rates per sex and one-year intervals. Data from the GBD studies with Dismod II (free at https://www.epigear.com/index_files/dismod_ii.html) can be used to derive internally consistent data and generate missing data. For example, the GBD studies provide data for incidence, prevalence and disease mortality, however, not case fatality. Other national level sources may also be explored/used, and compare with estimates produce from GBD data and Dismod II.

@@

Within Dismod II, each setting (e.g. country) has a collection that consists of population numbers (preferably the same as used in GBD studies, due to the mortality envelop) and all- cause mortality rates (numbers and calculate rates). The GBD provides 5-year age groups that are acceptable input parameters for Dismod II.

@@

Disability weights (DW) can be derived from disease specific years lived with disability (YLD) and disease specific prevalence by age group (5 years) and sex. Data for YLDs prevalence can be obtained from the online GBD data tool (http://ghdx.healthdata.org/gbd-results-tool). Our calculations of DW in the example here based on the GBD methods for estimating YLDs as the sum of sequelae prevalence multiplied by sequelae disability weights (REF GBD). The GBD has publicly available data at the cause level (e.g. ischemic heart disease) instead of squelae level (e.g. myocardial infarction, angina and heart failure). However, the GBD disability weights are for health states associated with sequelae, hence, we need to calculate DWs. An age and sex specific-correction was introduced to counteract the effects of accumulating comorbid illnesses in the older age groups (Equation 1).

@@

kable(pmslt_inputs,"latex", booktabs = T, caption = "PMSLT inputs") %>%
 kable_styling(latex_options = c("striped", "hold position"),
               full_width = F)%>%
  column_spec(2, width = "15em") %>%
  column_spec(3, width = "15em") %>%
  row_spec(0, bold = T)
```
\pagebreak

@@

We start by *cleaning the global environment* to keep track of our works and ensure that the code is generating our outcomes. Then, we set up an option to avoind the use of scientific notation and lastly we load the functions. The code chunks are shown in the rmarkdown ouput.

```{r, echo=TRUE}
rm (list = ls())

# Prevent scientific notation in data frame

options(scipen=999)

#Load all functions/packages

source("code/functions.R")
```

The data used here is for an example for Australia, however, this will be updated to link with the ITHIM-R modules (synthetic population, PIF calculations).

```{r, echo=TRUE}
##### Read data. idata: life table and disease life tables (including trends). edata: exposure data (e.g. physical activity).
#### irr.data: relative risks data and ee: energy expenditure
#### For exposure (edata) and input (idata) data I am using already generated categories, but a function should be developed to work from raw data.

idata <- read.csv("data/idata.csv", stringsAsFactors = F)
edata <- read.csv("data/edata.csv", stringsAsFactors = F)
irr <- read.csv("data/irr.csv", stringsAsFactors = F)
ee <- read.csv("data/ee.csv", stringsAsFactors = F)
```

>>>>

---
title: "Proportional multi-state multiple-cohort life table model"
author: "Belen Zapata-Diomedi and Ali Abbas"
date: "26 March 2018"
header-includes:
    - \usepackage{setspace}\onehalfspacing
    - \usepackage{float}
output:
   pdf_document: 
     toc: true
     toc_depth: 3
     number_sections: true
     fig_caption: yes
     fig_width: 7
     fig_height: 6
     df_print: kable
   fontsize: 12 pt
   geometry: margin=1in
bibliography: Bibliography.bib
---

@@

![Basic ITHIMR infrastructure](structure\Figure1.png)

@@

![Proportional multi-state life-table simplified framework. *The simplied PMST shows the interaction between the life table, disease life table and potential impact fraction (PIF). The PIF calculations by age and sex group are the same as those generated for ITHIM. The PIF (or 1-PIF) modifies incidence of disease, which changes prevalence and mortality (disease specific life table). Changes in prevalence and mortality rates from the disease specific life tables feed into the life table by changing all-cause mortality, which in turn changes life years. Change in prevalence of diseases changes total years lived with disability, which in turn modifies health-adjusted life years*](structure\Figure2.png)

@@

### Life table

Inputs of the life table are: population numbers by sex (per 1-year or age grouping of interest), mortality rates or probability of all cause mortality by single age group and sex and total prevalent years lived with disability rate per one year by sex. Disease specific disability weights are presented as inputs here as these adjust the total years lived with disability, hence, the health-adjusted life years.

@@

Data source: (1) National census; (2) Worldwide population and mortality data:  http://www.mortality.org/ (mostly high income countries; and (3) Calculate from the Global Buden of Disease by the Institute of Health Metrics and Evaluation (GBD IHME) data (rates and numbers available from (http://ghdx.healthdata.org/gbd-results-tool).

@@

Mortality rates are needed per single year and sex. These data are available from GBD IHME, however, in age groups (1-4, 5-9, etc). Interpolation can be used to derive in between ages rates (cubic spline).

Note that we need data for population numbers and all cause mortality rates for: (1) PMSLT and (2) Dismod II collection (more in Dismod II section). Population data from the synthetic population is used for the PMSLT. For Dismod II, popuation and mortality data should be from the same source (GBD IHME)

@@

Disability weights (DW) can be derived from disease specific years lived with disability (YLD) and disease specific prevalence by age group (5 years) and sex. Data for YLDs prevalence can be obtained from the online GBD IHME data tool (http://ghdx.healthdata.org/gbd-results-tool). Our calculations of DW in the example here based on the GBD methods for estimating YLDs as the sum of sequelae prevalence multiplied by sequelae disability weights (REF GBD). The GBD has publicly available data at the cause level (e.g. ischemic heart disease) instead of squelae level (e.g. myocardial infarction, angina and heart failure). However, the GBD disability weights are for health states associated with sequelae, hence, we need to calculate DWs. An age and sex specific-correction was introduced to counteract the effects of accumulating comorbid illnesses in the older age groups (Equation 1).

### Disease life tables

@@

For each of the modeled diseases the PMSLT needs incidence and case fatality rates per sex and one-year intervals. Data from the GBD IHME studies with Dismod II (free at https://www.epigear.com/index_files/dismod_ii.html) can be used to derive internally consistent data and generate missing data. For example, the GBD studies provide data for incidence, prevalence and disease mortality, however, not case fatality. Other national level sources may also be explored/used, and compare with estimates produce from GBD data and Dismod II.

@@

Within Dismod II, each setting (e.g. country) has a collection that consists of population numbers (preferably the same as used in GBD IHME studies, due to the mortality envelop) and all- cause mortality rates (numbers and calculate rates). The GBD provides 5-year age groups that are acceptable input parameters for Dismod II.

@@

@@

kable(pmslt_inputs,"latex", booktabs = T, caption = "PMSLT inputs") %>%
 kable_styling(latex_options = c("striped", "hold position"),
               full_width = F)%>%
  column_spec(2, width = "15em") %>%
  column_spec(3, width = "15em") %>%
  row_spec(0, bold = T)
```

@@

We start by cleaning the global environment (#1) to keep track of our works and ensure that the code is generating our outcomes. Then, we set up an option to avoid the use of scientific notation (#2) and lastly we load the functions (#3). The code chunks are shown in the rmarkdown ouput.

```{r, echo=TRUE}
#1
rm (list = ls())
#2
options(scipen=999)
#3
source("code/functions.R")
```

###Inputs

Table 1 describes data needs for the PMSLT, here we expand on the data needs and mechanisms (Figure 3) to use the PMSLT approach in ITHIMR (Figure 1). Figure 3 is used as a reference in the development of the code.

![Proportional multi-state life-table model. *Three sections are presented in Figure 3: **Data input sources**, **Inputs PMSLT** and **PMSLT mechanisms**.* ](structure\Figure3.png)

####Data for example

The code below is to load the data used in the example provided in this code for Australia (#4). *idata* includes data by one-year age groups and sex for: population numbers, mortality rates, total years lived with disability rates, incidence, case fatality, disability weights and costs for included diseases (ischemic heart disease, ischemic stroke, diabates, breast cancer (women) and colon cancer), cost for all other diseases (excludes those for physical inactivity) and trend for incidence and case fatality. Here we do not use cost and trends data. *edata* is for physical activity prevalence per categories (health risk factor), *irr* are relative risks and **ee** is energy expenditure per physical activity categories.

```{r, echo=TRUE}
#4
idata <- read.csv("data/idata.csv", stringsAsFactors = F)
edata <- read.csv("data/edata.csv", stringsAsFactors = F)
irr <- read.csv("data/irr.csv", stringsAsFactors = F)
ee <- read.csv("data/ee.csv", stringsAsFactors = F)
```

####Model parameters

We define parameters for our model that are used multiple times. *p_age_cohort* defines the starting age for the cohorts. *p_sex* is for the cohort sex. *p_disease* specifies the modelled diseases. *p_intervention* is a hypothetical increase in MET minutes per week.

```{r, echo=TRUE}
#5
p_age_cohort <- c(22, 27, 32, 37, 42, 47, 52, 57, 62, 67, 72, 77, 82, 87, 92, 97)
p_sex <- c("males", "females")
p_disease <- c("ihd", "istroke", "diabetes", "colon_cancer", "breast_cancer")
p_intervention_effect <- 100
```
